---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NegBinomHMC

<!-- badges: start -->

<!-- badges: end -->

NegBinomHMC is an R package that implements Hamiltonian Monte Carlo (HMC) sampling for Bayesian Negative Binomial regression models. The package provides functions to simulate Negative Binomial data, calculate log-posterior densities and gradients, and perform adaptive HMC sampling with a warmup phase to tune the step size automatically.

## Overview

Negative Binomial regression is commonly used for modeling over-dispersed count data. With NegBinomHMC, you can:

-   **Simulate Data:** Generate synthetic datasets following a Negative Binomial distribution using a specified regression model.

-   **Model Evaluation:** Calculate the log-posterior and its gradient for your Negative Binomial regression model.

-   **Adaptive HMC Sampling:** Sample from the Bayesian posterior using an adaptive HMC sampler that tunes its step size during a warmup period to achieve a desired acceptance rate.

-   **Overdispersion Parameter Estimation:** Jointly estimate the overdispersion parameter `r` using a Gamma prior.

This package is intended for educational purposes and as a starting point for more advanced Bayesian inference in count data models.

## Package Structure

``` bash
NegBinomHMC/
├── DESCRIPTION       # Package metadata
├── NAMESPACE         # Exports information generated by roxygen2
├── R/
│   ├── simulation.R       # Data simulation function
│   ├── negbin_model.R             # Negative Binomial model function
│   ├── hmc_sampler.R                # Adaptive HMC sampler and helper functions
├── man/              # Documentation files (generated via roxygen2)
├── README.md         # This file
└── LICENSE           # License file
```

## Installation

You can install the development version of NegBinomHMC from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Kumachar/NegBinomHMC")
```
And we strongly suggest users to install and run this package on Greatlakes, biostat cluster, or any local R environment with at least 16GB of RAM to avoid any possible issues.

## New Features: Dual Averaging for Adaptive Epsilon

### Concept

Dual averaging is a robust method to adapt `epsilon` by targeting a desired acceptance rate (denoted $\delta$, e.g., 0.65). It updates `epsilon` iteratively to minimize the difference between the actual and target acceptance probabilities. The algorithm:

1.  Runs HMC iterations, computing the acceptance probability $\alpha_t$ at iteration $t$.
2.  Updates a running statistic to adjust $\log \epsilon$.
3.  Smooths the updates to stabilize convergence.
4.  Use the final adapted `epsilon` for the sampling phase.

The update rules (from Hoffman & Gelman, 2014, used in NUTS) are:

$$
H_t = \left(1 - \frac{1}{t + t_0}\right) H_{t-1} + \frac{1}{t + t_0} (\delta - \alpha_t),
$$

$$
\log \tilde{\epsilon}_t = \mu - \frac{\sqrt{t}}{\gamma} H_t,
$$

and,

$$
\log \epsilon_t = \kappa \log \tilde{\epsilon}_t + (1 - \kappa) \log \epsilon_{t-1},
$$

where:

-   $\alpha_t$: Acceptance probability at iteration $t$.
-   $\delta$: Target acceptance rate (e.g., 0.65).
-   $H_t$: Running statistic tracking the error in acceptance rate.
-   $\mu$: Initial guess for $\log \epsilon$.
-   $t_0$: Controls early iteration weighting (e.g., 10).
-   $\gamma$: Controls adaptation speed (e.g., 0.05).
-   $\kappa$: Smoothing parameter (e.g., 0.75).
-   $\epsilon_t$: Step size at iteration $t$.

After a warm-up phase, we fix `epsilon` to the final adapted value for the sampling phase to ensure Markov chain validity. This means the algorithm may eventually have a different acceptance rate. But users can still use this feature to alter acceptance rate by set different accept rate r.

## Simulation Example

Below is a basic example demonstrating how to simulate Negative Binomials data and run the adaptive HMC sampler to estimate the regression coefficients. You can play with example by altering the params. To contrast the result we used a package called `MCMCpack` to run a random walk Metropolis sampler. It provides a function called `MCMCnegbin` to fit a Negative Binomial regression model using a auxiliary mixture sampling algorithm. Unlike our approach, it estimates the dispersion parameter `r` based on the outcome of generalized linear model (GLM). `MCMCnegbin` fails to provide valid estimates of `r` when the number of observations is less than 200 (not converge) or over 2000 (takes too many memories), while our HMC sampler can handle a larger range of observations. Here we set the number of observations to 500, which is a good trade-off between speed and accuracy. You can also try to increase the number of observations to 2000 and see the difference of these two methods.

```{r Simulation, warning=FALSE}
library(NegBinomHMC)
library(coda)
set.seed(123)

# Set simulation parameters
n <- 500              # Number of observations
p <- 3                # Number of predictors (including intercept)
beta_true <- c(3, 5, -3)  # True regression coefficients
r_true <- 2           # True dispersion parameter

# Simulate Negative Binomial data
data <- simulate_negbin_data(n, p, beta_true, r_true, seed = 123)
X <- data$X
y <- data$y
df <- data.frame(
  y  = y,
  x1 = X[, 2],          # drop the intercept column
  x2 = X[, 3]
)

# Prior hyperparameters
beta_mu <- rep(0, p)
beta_sigma <- rep(1, p)
alpha_r <- 5    # Stronger prior to avoid small r
beta_r <- 2     # Prior mean = 5/2 = 2.5, variance = 5/4 = 1.25

# Run the adaptive HMC sampler
initial_beta <- rep(0, p)  # Initial beta coefficients
initial_beta[3] <- -1
initial_theta <- c(initial_beta, log(2))  # Initial beta, phi = log(2)
start_time <- Sys.time()
result <- hmc_sampler(
  log_post = log_posterior_negbin_log_r,
  grad_log_post = grad_log_posterior_negbin_log_r,
  initial_theta = initial_theta,
  initial_epsilon = 0.005,  # Smaller step size
  L = 20,                   # More leapfrog steps
  n_iter = 10000,
  n_warmup = 2000,
  X = X,
  y = y,
  beta_mu = beta_mu,
  beta_sigma = beta_sigma,
  alpha_r = alpha_r,
  beta_r = beta_r,
  target_accept = 0.8
)
end_time <- Sys.time()
print(paste("HMC runtime:", end_time - start_time))

# Transform phi samples back to r for reporting
r_samples <- exp(result$samples[, p + 1])

# Results
cat("True beta:", beta_true, "\n")
cat("True r:", r_true, "\n")
cat("Estimated beta (mean):", colMeans(result$samples[, 1:p]), "\n")
cat("Estimated r (mean):", mean(r_samples), "\n")
cat("Acceptance rate:", result$acceptance_rate, "\n")
cat("Final epsilon:", result$final_epsilon, "\n")

# Posterior summaries
posterior_samples <- cbind(result$samples[, 1:p], r_samples)
colnames(posterior_samples) <- c(paste0("beta_", 1:p), "r")
summary(posterior_samples)

# Effective sample size
mcmc_chain <- as.mcmc(posterior_samples)
ess <- effectiveSize(mcmc_chain)
print("Effective Sample Size:")
print(ess)

# Length of 95% CI
ci_length <- apply(posterior_samples, 2, function(x) {
  quantile(x, 0.975) - quantile(x, 0.025)
})
print("Length of 95% CI for each parameter:")
print(ci_length)


par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
# Trace plot
for (j in 1:p) {
  plot(result$samples[, j], type = "l", 
       main = paste("Trace plot for beta", j),
       xlab = "Iteration",
       ylab = "Value")
  abline(h = beta_true[j], col = "red", lwd = 2)
}

# Density plot
for (j in 1:p) {
  plot(density(result$samples[, j]), 
       main = paste("Density plot for beta", j),
       xlab = "Value",
       ylab = "Density")
  abline(v = beta_true[j], col = "red", lwd = 2)
}
par(mfrow = c(2, 3))
plot(r_samples, type = "l", 
     main = "Trace plot for r", ylab = "Value", xlab = "Iteration")
abline(h = r_true, col = "red", lwd = 2)
plot(density(r_samples), 
     main = "Density plot for r",
     xlab = "Value",
     ylab = "Density")
abline(v = r_true, col = "red", lwd = 2)

par(mfrow = c(1, 1))

```

```{r MCMCnegbin simulation, warning=FALSE}
library(MCMCpack)
nb_draws <- MCMCnegbin(y ~ x1 + x2,          
                       data   = df,
                       burnin = 5e3,
                       mcmc   = 2e4,
                       thin   = 1,
                       seed   = 123)

summary(nb_draws)       # β’s and ρ (size) are already separated

# Effective Sample Size
mcmc_chain <- as.mcmc(nb_draws)
ess <- effectiveSize(mcmc_chain)
print("Effective Sample Size:")
print(ess)

nb_draws1 <- as.data.frame(nb_draws)
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
# Trace plot
for (j in 1:p) {
  plot(nb_draws1[, j], type = "l", 
       main = paste("Trace plot for beta", j),
       xlab = "Iteration",
       ylab = "Value")
  abline(h = beta_true[j], col = "red", lwd = 2)
}

# Density plot
for (j in 1:p) {
  plot(density(nb_draws1[, j]), 
       main = paste("Density plot for beta", j),
       xlab = "Value",
       ylab = "Density")
  abline(v = beta_true[j], col = "red", lwd = 2)
}

```

## Real-World Example 1

In this analysis, we investigate the factors influencing the number of days of absence among high school juniors. The data come from a study involving school administrators, where the predictors include the type of program in which a student is enrolled and a standardized math test score. Due to possible overdispersion in the count data, a Negative Binomial model is appropriate compared to a Poisson model.

We use our HMC package (NegBinomHMC) to sample from the posterior distribution of a Negative Binomial regression model. This document details the steps for reading in the data, building the design matrix, running the HMC sampler, and comparing the results with a baseline model.

### Data Reading and Preprocessing

We start by reading the dataset using the `haven` package. The dataset is stored in a Stata file and contains variables such as `math` (standardized test score), `prog` (program type), and `daysabs` (number of absence days). We convert the program type to a factor variable for modeling.

```{r real-world-example-1, warning=FALSE}
library(haven)
library(NegBinomHMC)
library(MCMCpack)
library(MASS)
library(ggplot2)

set.seed(123)
dat <- read_dta("https://stats.idre.ucla.edu/stat/stata/dae/nb_data.dta")
dat <- within(dat, {
  prog <- factor(prog, levels = 1:3,
                 labels = c("General","Academic","Vocational"))
  id <- factor(id)
})

# Design matrix and response
X <- model.matrix(~ math + prog, data = dat)
y <- dat$daysabs
p <- ncol(X)

# 1) Frequentist negative‑binomial
glm_fit <- glm.nb(daysabs ~ math + prog, data = dat)

# 2) HMC sampling
hmc_res <- hmc_sampler(
  log_post        = log_posterior_negbin_log_r,
  grad_log_post   = grad_log_posterior_negbin_log_r,
  initial_theta   = c(rep(0, p), log(10)),  # β=0, φ=log(10)
  initial_epsilon = 0.01,
  L               = 10,
  n_iter          = 10000,
  n_warmup        = 5000,
  X               = X, y = y,
  beta_mu         = 0,
  beta_sigma      = 1,
  alpha_r         = 5,
  beta_r          = 2,
  target_accept   = 0.8
)
hmc_samps <- hmc_res$samples
r_hmc     <- exp(hmc_samps[, p + 1])

# Acceptance rate
accept_rate <- hmc_res$acceptance_rate
cat("Acceptance rate:", accept_rate, "\n")

# 3) Random‑walk Metropolis
MCMCnegbin_chain <- MCMCnegbin(
  formula = daysabs ~ math + prog,
  data    = dat,
  burnin  = 0,
  mcmc    = 10000,
  tune    = 1,
  b0      = rep(0, p),
  B0      = diag(1, p),  # prior precision = 1 / 1²
  a0      = 5,
  c0      = 2,
  verbose = 0
)

# 4) Summaries: means & 95% CIs
params <- c(colnames(X), "r")

# HMC
hmc_means    <- c(colMeans(hmc_samps[, 1:p]),      mean(r_hmc))
hmc_ci_lower <- c(apply(hmc_samps[, 1:p], 2, quantile, probs = .025),
                   quantile(r_hmc, .025))
hmc_ci_upper <- c(apply(hmc_samps[, 1:p], 2, quantile, probs = .975),
                   quantile(r_hmc, .975))
names(hmc_means) <- names(hmc_ci_lower) <- names(hmc_ci_upper) <- params

# GLM
glm_means <- c(coef(glm_fit), r = glm_fit$theta)
conf_glm  <- confint(glm_fit)
glm_ci_lower <- glm_ci_upper <- rep(NA, length(params))
names(glm_ci_lower) <- names(glm_ci_upper) <- params
for(nm in rownames(conf_glm)) {
  glm_ci_lower[nm] <- conf_glm[nm, 1]
  glm_ci_upper[nm] <- conf_glm[nm, 2]
}

# MCMCnegbin
MCMCnegbin_means    <- colMeans(MCMCnegbin_chain)
MCMCnegbin_ci_lower <- apply(MCMCnegbin_chain, 2, quantile, probs = .025)
MCMCnegbin_ci_upper <- apply(MCMCnegbin_chain, 2, quantile, probs = .975)

# 5) Build long data.frame for β’s
df_beta <- data.frame(
  Parameter = rep(colnames(X), 3),
  Method    = factor(rep(c("HMC","GLM","MCMCnegbin"), each = p),
                     levels = c("HMC","GLM","MCMCnegbin")),
  Estimate  = c(hmc_means[colnames(X)],
                glm_means[colnames(X)],
                MCMCnegbin_means[colnames(X)]),
  CI_Lower  = c(hmc_ci_lower[colnames(X)],
                glm_ci_lower[colnames(X)],
                MCMCnegbin_ci_lower[colnames(X)]),
  CI_Upper  = c(hmc_ci_upper[colnames(X)],
                glm_ci_upper[colnames(X)],
                MCMCnegbin_ci_upper[colnames(X)])
)
df_beta$Parameter <- factor(df_beta$Parameter, levels = colnames(X))

df_beta$CI_length <- df_beta$CI_Upper - df_beta$CI_Lower

# 6) Plot β’s with 95% CIs, add length of CIs
ggplot(df_beta, aes(x = Parameter, y = Estimate, color = Method)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper),
                position = position_dodge(width = 0.6), width = 0.2) +
  # label the CI length
  geom_text(aes(label = round(CI_length, 2),    # round as needed
                y     = CI_Upper),              # position text at top of bar
            position = position_dodge(width = 0.6),
            vjust    = -0.5,                    # nudge it a bit higher
            size     = 3) +                     # text size
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "β Estimates with 95% CIs\n(Length of CI shown above)",
       x     = NULL,
       y     = "Estimate")

# 7) Plot dispersion “r”
df_r <- data.frame(
  Parameter = factor("r"),
  Method    = factor(c("HMC","GLM","MCMCnegbin"),
                     levels = c("HMC","GLM","MCMCnegbin")),
  Estimate  = c(hmc_means["r"], glm_means["r"], MCMCnegbin_means["r"]),
  CI_Lower  = c(hmc_ci_lower["r"], glm_ci_lower["r"], MCMCnegbin_ci_lower["r"]),
  CI_Upper  = c(hmc_ci_upper["r"], glm_ci_upper["r"], MCMCnegbin_ci_upper["r"])
)

print("df_r:")
print(df_r)

ggplot(df_r, aes(x = Parameter, y = Estimate, color = Method)) +
  geom_point(position = position_dodge(width = 0.3), size = 4) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper),
                position = position_dodge(width = 0.3), width = 0.1) +
  theme_bw() +
  labs(title = "Dispersion (r) Estimates with 95% CIs", x = NULL, y = "Value")



# HMC trace & density
par(mfrow = c(2, 4), mar = c(4, 4, 2, 1))
coln <- colnames(X)
for(pn in 1:4) {
  plot(hmc_samps[,pn], type="l", main=paste("HMC Trace:", coln[pn]),
       xlab="Iteration", ylab="Value")
  abline(h = hmc_means[pn], col = "red", lty = 2)
}
for(pn in 1:4) {
  d <- density(hmc_samps[,pn])
  plot(d, main=paste("HMC Density:", coln[pn]),
       xlab="Value", ylab="Density")
  abline(v = hmc_means[pn], col = "red", lty = 2)
}
par(mfrow = c(1, 1))

# MCMCnegbin trace
MCMCnegbin_chain <- as.data.frame(MCMCnegbin_chain)
par(mfrow = c(2, 4), mar = c(4, 4, 2, 1))

for(pn in 1:4) {
  plot(MCMCnegbin_chain[,pn], type="l", main=paste("MCMCnegbin Trace:", coln[pn]),
       xlab="Iteration", ylab="Value")
  abline(h = MCMCnegbin_means[pn], col = "blue", lty = 2)
}
for(pn in 1:4) {
  d <- density(MCMCnegbin_chain[,pn])
  plot(d, main=paste("MCMCnegbin Density:", coln[pn]),
       xlab="Value", ylab="Density")
  abline(v = MCMCnegbin_means[pn], col = "blue", lty = 2)
}
par(mfrow = c(1, 1))

```

## Real-World Example 2

In this analysis, we investigate the factors influencing the total number of injuries in traffic accidents. The data come from a traffic accidents dataset and include predictors such as lighting condition, number of units involved, day of the week, and month of the crash. Considering possible overdispersion in the count data, we use a Negative Binomial regression model. This document demonstrates how to fit the model using both a traditional approach (via `glm.nb` from the MASS package) and our custom Hamiltonian Monte Carlo (HMC) sampler.

### Data Preprocessing

We start by reading in the `traffic_accidents.csv` file. The categorical variables are converted to factors, and numerical values (e.g., number of units) are properly set as numeric.


```{r real-world-example-2, warning=FALSE}
library(NegBinomHMC)
library(MCMCpack)
library(MASS)
library(ggplot2)

accidents <- read.csv("ExampleData/traffic_accidents.csv", stringsAsFactors = TRUE)

accidents$weather_condition <- as.factor(accidents$weather_condition)
accidents$lighting_condition <- as.factor(accidents$lighting_condition)
accidents$crash_day_of_week <- as.factor(accidents$crash_day_of_week)
accidents$crash_month <- as.factor(accidents$crash_month)
accidents$num_units <- as.numeric(as.character(accidents$num_units))

nb_model <- glm.nb(injuries_total ~ lighting_condition + num_units + crash_day_of_week + crash_month,
                   data = accidents)
summary(nb_model)

nb_formula <- injuries_total ~ lighting_condition + num_units + crash_day_of_week + crash_month
X <- model.matrix(nb_formula, data = accidents)
y <- accidents$injuries_total

beta_mu <- rep(0, ncol(X))
beta_sigma <- rep(10, ncol(X))

initial_beta <- rep(0, ncol(X))

initial_epsilon <- 0.1 
L <- 10 
n_iter <- 1000
n_warmup <- 500
target_accept <- 0.65
alpha_r <- 5    # Gamma prior shape for r
beta_r <- 2     # Gamma prior rate for r, mean = 5/2 = 2.5

hmc_result <- hmc_sampler(
  log_post = log_posterior_negbin_log_r,
  grad_log_post = grad_log_posterior_negbin_log_r,
  initial_theta = c(initial_beta, log(0.74)),  # Initial beta = 0, phi = log(0.74)
  initial_epsilon = initial_epsilon,
  L = L,
  n_iter = n_iter,
  n_warmup = n_warmup,
  X = X,
  y = y,
  beta_mu = beta_mu,
  beta_sigma = beta_sigma,
  alpha_r = alpha_r,
  beta_r = beta_r,
  target_accept = target_accept
)

samples <- hmc_result$samples
accept_rate <- hmc_result$acceptance_rate
final_epsilon <- hmc_result$final_epsilon

cat("HMC acceptance rate:", accept_rate, "\n")
cat("final epsilon:", final_epsilon, "\n")

# Transform phi samples to r
r_samples <- exp(samples[, ncol(X) + 1])
posterior_means <- c(colMeans(samples[, 1:ncol(X)]), mean(r_samples))
names(posterior_means) <- c(colnames(X), "r")
cat("HMC posterior mean:\n")
print(posterior_means)

cat("glm model estimated coefficients:\n")
print(c(coef(nb_model), theta = nb_model$theta))

ci_length <- apply(cbind(samples[, 1:ncol(X)], r_samples), 2, function(x) {
  quantile(x, 0.975) - quantile(x, 0.025)
})
print("Length of 95% CI for each parameter:")
print(ci_length)

# Metropolis model using MCMCnegbin
iterations <- 1000
chain_metropolis <- MCMCnegbin(
  formula = nb_formula,
  data = accidents,
  burnin = 0,  # No burn-in
  mcmc = iterations - 1,  # Include initial value
  tune = 1,  # Tuning parameter for proposal
  b0 = beta_mu,  # Prior mean for beta
  B0 = diag(1/beta_sigma^2, ncol(X)),  # Prior precision for beta
  a0 = alpha_r,  # Gamma prior shape for r
  c0 = beta_r,   # Gamma prior rate for r
  verbose = 0
)

cat("Random Walk Metropolis acceptance rate:", 1 - mean(rowSums(chain_metropolis[-1, ] == chain_metropolis[-nrow(chain_metropolis), ]) == ncol(chain_metropolis)), "\n")
MCMCnegbin_posterior_means <- colMeans(chain_metropolis)
names(MCMCnegbin_posterior_means) <- colnames(X)
cat("Random Walk Metropolis posterior mean:\n")
print(MCMCnegbin_posterior_means)

ci_length <- apply(chain_metropolis, 2, function(x) {
  quantile(x, 0.975) - quantile(x, 0.025)
})
print("Length of 95% CI for each parameter (MCMCnegbin, beta only):")
print(ci_length)

# ===== Compute HMC estimates and CI =====
hmc_mean <- posterior_means
hmc_ci_lower <- apply(cbind(samples[, 1:ncol(X)], r_samples), 2, quantile, probs = 0.025)
hmc_ci_upper <- apply(cbind(samples[, 1:ncol(X)], r_samples), 2, quantile, probs = 0.975)

# ===== Compute MCMCnegbin estimates and CI =====
MCMCnegbin_mean <- MCMCnegbin_posterior_means
MCMCnegbin_ci_lower <- apply(chain_metropolis, 2, quantile, probs = 0.025)
MCMCnegbin_ci_upper <- apply(chain_metropolis, 2, quantile, probs = 0.975)

# ===== Compute GLM estimates and CI =====
glm_coef <- c(coef(nb_model), theta = nb_model$theta)
# Align glm_coef with params
params <- c(colnames(X), "r")
glm_coef_aligned <- numeric(length(params))
names(glm_coef_aligned) <- params
glm_coef_names <- names(coef(nb_model))
for (name in glm_coef_names) {
  if (name %in% colnames(X)) glm_coef_aligned[name] <- coef(nb_model)[name]
}
glm_coef_aligned["r"] <- nb_model$theta

# Create glm_ci with NA for missing coefficients
glm_ci <- matrix(NA, nrow = length(params), ncol = 2)
rownames(glm_ci) <- params
colnames(glm_ci) <- c("2.5%", "97.5%")
confint_glm <- confint(nb_model)
for (name in rownames(confint_glm)) {
  if (name %in% colnames(X)) glm_ci[name, ] <- confint_glm[name, ]
}
glm_ci["r", ] <- c(NA, NA)


# Diagnostic prints to check dimensions
cat("Number of parameters (p + 1):", length(params), "\n")
cat("Length of hmc_mean:", length(hmc_mean), "\n")
cat("Length of glm_coef_aligned:", length(glm_coef_aligned), "\n")
cat("Length of MCMCnegbin_mean:", length(MCMCnegbin_mean), "\n")
cat("Number of rows in glm_ci:", nrow(glm_ci), "\n")

# ===== Prepare a data frame for plotting beta parameters =====
df_beta <- data.frame(
  Parameter = rep(params[1:24], 3),
  Method = factor(rep(c("HMC", "GLM", "MCMCnegbin"), each = 24)),
  Estimate = c(hmc_mean[1:24], glm_coef_aligned[1:24], MCMCnegbin_mean),
  CI_Lower = c(hmc_ci_lower[1:24], glm_ci[1:24, 1], MCMCnegbin_ci_lower),
  CI_Upper = c(hmc_ci_upper[1:24], glm_ci[1:24, 2], MCMCnegbin_ci_upper)
)

df_beta$Parameter <- factor(df_beta$Parameter, levels = params[1:24])

# ===== Create the plot for beta parameters =====
ggplot(df_beta, aes(x = Parameter, y = Estimate, color = Method)) +
  geom_point(position = position_dodge(width = 0.6), size = 3) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper),
                width = 0.2,
                position = position_dodge(width = 0.6)) +
  theme_bw() +
  labs(title = "Beta Parameter Estimates with 95% Intervals",
       x = "Parameter",
       y = "Coefficient Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ===== Prepare a data frame for plotting r (HMC and GLM only) =====
df_r <- data.frame(
  Parameter = rep("r", 2),
  Method = factor(c("HMC", "GLM")),
  Estimate = c(hmc_mean["r"], glm_coef_aligned["r"]),
  CI_Lower = c(hmc_ci_lower["r_samples"], glm_ci["r", 1]),
  CI_Upper = c(hmc_ci_upper["r"], glm_ci["r", 2])
)

print("df_r:")
print(df_r)

# ===== Create the plot for r =====
ggplot(df_r, aes(x = Parameter, y = Estimate, color = Method)) +
  geom_point(position = position_dodge(width = 0.2), size = 3) +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper),
                width = 0.1,
                position = position_dodge(width = 0.2)) +
  theme_bw() +
  labs(title = "Dispersion Parameter (r) Estimates with 95% Intervals",
       x = "Parameter",
       y = "Value") +
  theme(axis.text.x = element_text(hjust = 0.5))

first_three <- params[1:3]

colnames(samples) <- c(colnames(X), "r")
colnames(chain_metropolis) <- colnames(X)

# --- Set up the plotting area: 2 rows (first row for HMC, second row for MCMCnegbin), 3 columns ---
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))

# --- Generate trace plots for HMC and add mean line ---
for (i in 1:3) {
  param_name <- first_three[i]
  plot(samples[, param_name],
       type = "l",
       main = paste("HMC Trace:", param_name),
       xlab = "Iteration",
       ylab = "Value")
  
  hmc_mean_val <- mean(samples[, param_name])
  
  abline(h = hmc_mean_val, col = "red", lwd = 2, lty = 2)
  
  legend("topright", legend = paste("Mean:", round(hmc_mean_val, 3)), 
         bty = "n", col = "red", lty = 2, cex = 0.8)
}

# density plot
for (i in 1:3) {
  param_name <- first_three[i]
  plot(density(samples[, param_name]), 
       main = paste("HMC Density:", param_name),
       xlab = "Value",
       ylab = "Density")
  hmc_mean_val <- mean(samples[, param_name])
  abline(v = hmc_mean_val, col = "red", lwd = 2, lty = 2)
  
  legend("topright", legend = paste("Mean:", round(hmc_mean_val, 3)), 
         bty = "n", col = "red", lty = 2, cex = 0.8)
}

# --- Generate trace plots for MCMCnegbin and add mean line ---
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
chain_metropolis <- as.data.frame(chain_metropolis)
for (i in 1:3) {
  param_name <- first_three[i]
  plot(chain_metropolis[, param_name],
       type = "l",
       main = paste("MCMCnegbin Trace:", param_name),
       xlab = "Iteration",
       ylab = "Value")
  
  MCMCnegbin_mean_val <- mean(chain_metropolis[, param_name])
  
  abline(h = MCMCnegbin_mean_val, col = "blue", lwd = 2, lty = 2)
  
  legend("topright", legend = paste("Mean:", round(MCMCnegbin_mean_val, 3)), 
         bty = "n", col = "blue", lty = 2, cex = 0.8)
}

# density plot
for (i in 1:3) {
  param_name <- first_three[i]
  plot(density(chain_metropolis[, param_name]), 
       main = paste("MCMCnegbin Density:", param_name),
       xlab = "Value",
       ylab = "Density")
  
  MCMCnegbin_mean_val <- mean(chain_metropolis[, param_name])
  
  abline(v = MCMCnegbin_mean_val, col = "blue", lwd = 2, lty = 2)
  
  legend("topright", legend = paste("Mean:", round(MCMCnegbin_mean_val, 3)), 
         bty = "n", col = "blue", lty = 2, cex = 0.8)
}

# Reset the plotting layout to one plot per window
par(mfrow = c(1, 1))
```


## Contributing

Contributions to NegBinomHMC are welcome. If you encounter issues or have suggestions for improvements, please feel free to open an issue or submit a pull request on the GitHub repository.

## License

This project is licensed under the MIT License. See the LICENSE file for details.

## Contact

For questions or further information, please contact Weixiong at [[wxhua\@umich.edu](mailto:wxhua@umich.edu){.email}].
